Extended Data

### 2021-10-06 doc/manuscript/Dow_manuscript_Nature.Rmd
edit Dendrobands section

##figures
EDF_map = "Extended Data Figure 1" 
EDF_growth_curves_all = "Extended Data Figure 2"  
EDF_climwin = "Extended Data Figure 3"
EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
EDF_tree_rings_TMIN = "Extended Data Figure 7"

@@

While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few sites (e.g., Ref.[@dorangeville_peak_2021]). 
<!-- one example will suffice here. others include one cited in Anderson-Teixeira et al. GCB tree rings. martin_hydrometeorology_2017 does this, but that is for conifers.-->
To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

@@

Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).

@@

Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing`, `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the critical temperature window (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001). 
Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest *when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window*<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression. We want the latter, and to remove the italic text -->, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).
Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.

Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig. `r DOY_timing`), there was no consistent lengthening of $L_{PGS}$. 
Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`). 
Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig. `r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin` - `r EDF_pheno_Tmin_AprilMay`).
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
while $\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`). 
Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth.

>>>>

##figures
EDF_map = "Extended Data Figure 1" 
EDF_growth_curves_all = "Extended Data Figure 2"  
EDF_climwin = "Extended Data Figure 3"
EDF_pheno_all = "Extended Data Figures 4-6"
EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
EDF_tree_rings_TMIN = "Extended Data Figure 7"

@@

While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with *multi-year records of growth phenology having been published for only one site[@dorangeville_peak_2021]*. 
<!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
To our knowledge, the climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

@@

Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** <!--Cameron, please verify number--> trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** <!--Cameron, please verify number--> trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).

@@

Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-6), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the critical temperature window (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001). 
Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest ~~when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window~~<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression. We want the latter, and to remove the italic text -->, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.

Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig. `r DOY_timing`), there was no consistent lengthening of $L_{PGS}$. 
Specifically, because both $DOY_{25}$ and $DOY_{75}$ advanced in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, `r EDF_pheno_all`).

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to mean $T_{max}$ during the critical temperature window (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest).
$\Delta DBH$ displayed no relationship with mean $T_{max}$ during the critical temperature window (`r EDF_pheno_Tmax_climwin`). 
Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth.
